---
title: "Assignment 1"
subtitle: "Density Estimation and Clustering"
author: "Alexander, Ulrik, Hannes"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output: 
   html_document:
    code_folding: hide
    theme: cerulean
    highlight: textmate
    number_sections: false
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Boston Housing Data

```{r}
data("Boston", package = "MASS")
X <- scale(Boston[,c(13,6)])
plot(X,as=1)
```

# Questions

## 1. Estimate Joint Bivariate Density

We want to estimate the joint bivariate density using a kernel estimator with the same bandwidth in both dimensions: $h = (a, a)$. 

```{r}
library(sm)
plot(X,as=1,col=8)

# Select h. Not sure how we should use the hint. 
# Which of the following values should we trust?
# Looks like we should find this maximum manually, but how?
h1 <- h.select(X, method = "cv")
h2 <- hcv(X, method = "cv")

# Density estimation using the value of a found above. 
sm.density(X, h = h1[[1]]*c(1,1), display="slice",props=c(25,50,75,95),col=2,add=TRUE)
sm.density(X, h = 0.1*c(1,1), display="slice",props=c(25,50,75,95),col=2,add=TRUE)
```

The maximum log-likelihood cross-validation method will be used to choose the value of $a$. Then, the density will be estimated using the value of $a$.

```{r}
# Try to estimate "manually" to check if we get similar value to above. 
# Does not give sensible results. How should we maximize this manually? :))

lstat.min <- min(X[,1])
lstat.max <- max(X[,1])
rm.min <- min(X[,2])
rm.max <- max(X[,2])
grid.values <- 15
lst <- seq(from = lstat.min, to = lstat.max, length.out = grid.values)
rme <- seq(from = rm.min, to = rm.max, length.out = grid.values)
maxes <- c() # Slow, but alright for testing I guess. 
objectives <- c() # Slow, but alright for testing I guess.
for (i in 1:grid.values){
   for (j in 1:grid.values){
      new.point <- matrix(c(lst[i], rme[j]), ncol = 2)
      f <- function(x){
         f.hat <- sm.density(X,h=x*c(1,1),display="none",eval.grid=FALSE, eval.points = new.point)
         log(f.hat$estimate)
      }
      maxes <- c(maxes, optimize(f, interval = c(0.1, 1), maximum = T)$maximum)
      objectives <- c(objectives, optimize(f, interval = c(0.1, 1), maximum = T)$objective)
   }
}

m.id <- which.max(objectives)
maxes[m.id]

# The 1d-optimization in optimize depends on at which point the sm is evaluated. Want to try to maximize over the
# entire grid + maximize on the given range of discrete a-values. 
```

## 2. Hierarchical Clustering

```{r}
library(ggplot2)
library(dplyr)
hcl1 <- hclust(dist(X), method="ward.D")
plot(hcl1)
clusts <- cutree(hcl1, 3)

df <- data.frame(lstat = X[,1], rm = X[,2], col = factor(clusts))
ggplot(df, aes(x = rm, y = lstat)) + 
   geom_point(aes(colour = col)) +
   theme_minimal()

```


## 3. For each of the Clusters, do ...

### Cluster 1

```{r}
# Data from cluster 1.
clust1.data <- df %>% dplyr::filter(col == "1")

# Non-parametric estimation of the joint density. 
plot(clust1.data[, c(1,2)])
sm.density(clust1.data[, c(1,2)], h = h1[[1]]*c(1,1), display="slice",props=c(75),col=2,add=TRUE)
```

### Cluster 2

```{r}
# Data from cluster 1.
clust2.data <- df %>% dplyr::filter(col == "2")

# Non-parametric estimation of the joint density. 
plot(clust2.data[, c(1,2)])
sm.density(clust2.data[, c(1,2)], h = h1[[1]]*c(1,1), display="slice",props=c(75),col=2,add=TRUE)
```

### Cluster 3

```{r}
# Data from cluster 1.
clust3.data <- df %>% dplyr::filter(col == "3")

# Non-parametric estimation of the joint density. 
plot(clust3.data[, c(1,2)])
sm.density(clust3.data[, c(1,2)], h = h1[[1]]*c(1,1), display="slice",props=c(75),col=2,add=TRUE)
```

## 4. Repeat 2 and 3, with Automatically Selected Optimal Number of Clusters
