---
title: "Example of Local MDS"
author: "Pedro Delicado"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Portion of S shape curve, with radious R, of length 2*pi with noise

t <- seq(-1.5*pi,1.5*pi,l=100)
R<- 1
n<-75
sd.residuals <- .15
```

```{r, fig.asp=1}
# windows(width=5,height=5)
# pdf("ej_MDS_S.pdf",width=5,height=5,paper="special")
set.seed(1)
A <- 2.6
plot(0,0,type="n",xlab="",ylab="",axes=FALSE,
#     main="Ilustraci?n Paso 1 de ISOMAP",
     xlim=c(-1-2*sd.residuals,1+2*sd.residuals + A),
     ylim=c(-2-2*sd.residuals,2+2*sd.residuals))

# S plot with points
y <- R*sign(t) - R*sign(t)*cos(t/R)
x <- -R*sin(t/R)
lines(x,y,col=2,lwd=2)
rt <- sort(runif(n)*3*pi - 1.5*pi)
eps <- rnorm(n)*sd.residuals
ry <- R*sign(rt) - (R+eps)*sign(rt)*cos(rt/R)
rx <- -(R+eps)*sin(rt/R)
points(rx,ry,col=1,pch=19,cex=.6)
XY <- cbind(rx,ry)
#lines(x,y,col=2,lwd=2)

# S plot with points and all the connections
temp.seed <- .Random.seed
for (i in 1:(n-1)){
   for (j in (i+1):n){
       if ( runif(1) <= .1 ) lines(A + rx[c(i,j)],ry[c(i,j)],col=3)
   }
}
lines(A+x,y,col=2,lwd=2)
points(A + rx,ry,col=1,pch=19,cex=.6)
#dev.off()
```

```{r}
# windows(width=10,height=4)
# pdf("ej_MDS_S_res.pdf",width=10,height=4,paper="special")
mdsXY <- cmdscale(dist(XY),eig = TRUE)
#op<-par(mfrow=c(1,2))
plot(mdsXY$eig,type="b",main="MDS eigenvalues")
```

```{r}
plot(mdsXY$points%*%diag(c(1,-1)),main="MDS configuration")
#par(op)
#dev.off()
```

```{r}
#windows(width=7,height=5)
# pdf("ej_localMDS_S.pdf",width=7,height=5,paper="special")
set.seed(1)
A <- 2.6
plot(0,0,type="n",xlab="",ylab="",axes=FALSE,
     xlim=c(-1-2*sd.residuals,1+2*sd.residuals + 2*A),
     ylim=c(-2-2*sd.residuals,2+2*sd.residuals))

# S plot with points
y <- R*sign(t) - R*sign(t)*cos(t/R)
x <- -R*sin(t/R)
lines(x,y,col=2,lwd=2)
rt <- sort(runif(n)*3*pi - 1.5*pi)
eps <- rnorm(n)*sd.residuals
ry <- R*sign(rt) - (R+eps)*sign(rt)*cos(rt/R)
rx <- -(R+eps)*sin(rt/R)
points(rx,ry,col=1,pch=19,cex=.6)
#lines(x,y,col=2,lwd=2)

# S plot with points and all the connections
temp.seed <- .Random.seed
for (i in 1:(n-1)){
  for (j in (i+1):n){
    if ( runif(1) <= .1 ) lines(A + rx[c(i,j)],ry[c(i,j)],col=3)
  }
}
lines(A+x,y,col=2,lwd=2)
points(A + rx,ry,col=1,pch=19,cex=.6)

# S plot with points, with "short" and "large" connections
eps <- .75#.5
set.seed(temp.seed)
for (i in 1:(n-1)){
  for (j in (i+1):n){
    if ( runif(1) <= .1 ) lines(2*A + rx[c(i,j)],ry[c(i,j)],col=3)
  }
}
for (i in 1:(n-1)){
  for (j in (i+1):n){
    if ( dist(XY[c(i,j),])<= eps )
      lines(2*A + rx[c(i,j)],ry[c(i,j)],col=4,lwd=2)
  }
}
lines(2*A+x,y,col=2,lwd=2)
points(2*A + rx,ry,col=1,pch=19,cex=.6)
text(2*A,2.35,"Short and large distances")
#dev.off()
```

```{r}
# Removing large distances: smacof with weights
library(smacof)
distXY <- dist(XY)
n <- dim(XY)[1]

# pdf("ej_weighted_MDS_S_res.pdf",width=12,height=6,paper="special")
set.seed(1)
op<-par(mfrow=c(1,2))
k <- 5
w <- dissWeights(distXY, type = "knn", k = k)
w <- w * (n * (n - 1) * 0.5/sum(w^2))^0.5
res <- mds(distXY, weightmat = w, init = "random")
plot(res$conf,main=paste("0-1 weighted MDS, K=",k))
text(res$conf[,1],res$conf[,2],1:n,pos=3)

k <- 10
w <- dissWeights(distXY, type = "knn", k = k)
w <- w * (n * (n - 1) * 0.5/sum(w^2))^0.5
res <- mds(distXY, weightmat = w, init = "random")
plot(res$conf,main=paste("0-1 weighted MDS, K=",k))
text(res$conf[,1],res$conf[,2],1:n,pos=3)

par(op)
#dev.off()
```

```{r}
# Local MDS. Implemented by Pedro Delicado, December 2020
#
# Objective function in Local MDS, eq (7) in Chen and Buja (2009)
Stress.LocalMDS <- function(conf,n,dist.orig,k=5,tau=1){
  require(smacof)
  q<-length(conf)/n
  mconf <- matrix(conf,nrow = n, byrow = FALSE) # configuration matrix
  Eucl.dist <- dist(mconf)
  w <- dissWeights(dist.orig, type = "knn", k = k)
  w.1 <- (w==1)
  card.N.k <- sum(w.1)
  card.N.k.c <- n*(n-1)/2 - card.N.k
  t <- (card.N.k/card.N.k.c)*median(dist.orig[w.1])*tau
  return(sum((dist.orig[w.1]-Eucl.dist[w.1])^2) - t*sum(Eucl.dist[!w.1]))
}
```

```{r,fig.height=6}
k <- 5
tau <- .05
q<-2 # 2-dim config
conf0 <- as.numeric( cmdscale(distXY,k=q) )
localMDS.S.res <- optim(par=conf0, fn=Stress.LocalMDS, n=n,dist.orig=distXY,k=k,tau=tau,
                        method ="BFGS", control = list(maxit=1000))
conf.localMDS.S.res <- matrix(localMDS.S.res$par, nrow = n, byrow = FALSE)
# pdf("ej_Local_MDS_S_res.pdf",width=10,height=8,paper="special")
par(mfrow=c(2,1))
plot(conf.localMDS.S.res,as=1, main=paste0("Local MDS, k=",k,", tau=",tau))
text(conf.localMDS.S.res[,1],conf.localMDS.S.res[,2],1:n,pos=3)
#
plot(conf.localMDS.S.res[,1],rt)
par(op)
#dev.off()
```

It seems that $q=1$ can be enough:
```{r}
k <- 5
tau <- .05
q<-1 # 1-dim config
conf0 <- as.numeric(cmdscale(distXY,k=q))
localMDS.S.res <- optim(par=conf0, fn=Stress.LocalMDS,
                      n=n,dist.orig=distXY,k=k,tau=tau,
                      method ="BFGS", 
                      control = list(maxit=1000))
conf.localMDS.S.res <- matrix(localMDS.S.res$par, nrow = n, byrow = FALSE)
```

```{r}
lambda <- conf.localMDS.S.res[,1] #1-dimemsional configuration
plot(lambda,rep(0,n),#as=1,    
     main=paste0("Local MDS, k=",k,", tau=",tau),
     xlab="lambda (1-dimemsional configuration)",
     ylab="")
text(lambda,rep(0,n),1:n,pos=3,cex=.4)
```

The 1-dimensional configuration $\lambda$ given by Local MDS is a good recovery of the true underline 1-dimensional generating variable `rt`. The linear relation between $\lambda$ and `rt` is almost perfect:

```{r,fig.asp=1}
plot(lambda,rt, as=1)
```

Now we plot the 1-dimensional structure of the data obtained from Local MDS using smoothing techniques:

```{r,fig.height=8}
#install.packages("mgcv") # To perform spline smoothing
library(mgcv)
smooth.x <- gam(rx~s(lambda))
smooth.y <- gam(ry~s(lambda))

s.x <- smooth.x$fitted.values
s.y <- smooth.y$fitted.values

op <- par(mfrow=c(2,1))
plot(lambda,rx,col=8, pch=19, cex=1, main="x versus lambda")
lines(lambda, s.x, col=1, lwd=2)
plot(lambda,ry,col=8, pch=19, cex=1, main="y versus lambda")
lines(lambda, s.y, col=1, lwd=2)
par(op)
```
We add the nonlinear curve fitted by Local MDS to the 2-dimensional representation of the original data:

```{r}
A <- 2.6
plot(0,0,type="n",xlab="",ylab="",axes=FALSE,
     xlim=c(-1-2*sd.residuals,1+2*sd.residuals + 2*A),
     ylim=c(-2-2*sd.residuals,2+2*sd.residuals))

# S plot with points
lines(x,y,col=2,lwd=2)
points(rx,ry,col=8,pch=19,cex=.8)
lines(s.x,s.y,col=1,lwd=2)

legend(.75*A,.75*A,c("Observed points","True generating curve", "Estimated curve"),
       col=c(8,2,1), pch=c(19,NA,NA), lwd=c(NA,2,2), bty="n", cex=.8)
```
